Bee visitation, pollination services, and plant yield in commodity and hybrid seed canola

### 2022-07-19 Manuscript/main.tex
Worked on partial and SEM plots

Honey bee visitation decreased with distance away from the apiary, in both commodity and seed canola fields.
In commodity fields, honey bee visitation decreased from 6.4 visits/hr at the edge to 1.0 visits/hr at the centre of the field (400 m) (p$=$0.0002, Figure \ref{fig:hbeeDist_commodity}, apiary of 40 hives).
% Visitation rates were higher in both years in Lethbridge (2014: 9.3 visits/hr, 2015: 7.1), but were higher in 2015 than 2014 at Grande Prairie (Z=2.3, p$=$0.02), rising from 5 visits/hr at the field edge in 2014 to 17 visits/hr in 2015. % (at a stocking level of 20 hives). 
Higher stocking rates increased honey bee visitation at the edge of the field from 3.9 visits/hr with an apiary of 20 hives, increasing to 6.4 visits/hr with 40 hives (Z=5.3, p$<$0.0001), while unstocked field edges had an average of 0.4 visits/hr.
Honey bee visitation was much higher overall in seed canola fields, and decreased from 99.5 visits/hr at edge to 78.8 visits/hr at the centre of the field (Z=-2.5, p$=$0.01). 
Honey bee visitation was also lower near leafcutter shelters (Z=3.9, p$=$0.001), dropping from 60 visits/hr at 10 m to 34 visits/hr at 2 m, likely due to competition with LCBs (Figure \ref{fig:hbeeDist_both}, \ref{fig:lbeeDist_both}).

@@

\begin{figure}
    \centering
    \includegraphics[width=0.9\textwidth,keepaspectratio=true]{../Figures/allVisits.png}
    \caption{Effect of distance from a) honey bee apiaries and b) leafcutter shelters on visitation rate. Visitation rates in commodity fields are shown at a stocking rate of 40 hives (black line), while stocking rates are at 160 hives in seed fields (yellow line).}
    \label{fig:hbeeDist_commodity}
\end{figure}

@@

LCB visitation increased pollen deposition in seed canola fields, but honey bees did not increase pollen deposition in either commodity or seed canola fields.
Pollen deposition on stigmas was high in commodity canola (mean: 293 grains per stigma, SD: 385, range: 0--3981), but honey bee visitation had no impact on this (p$=$0.77, Figure \ref{fig:hbeeVisPol}).
Pollen deposition slighly declined with distance from the edge of the field, but this relationship was weak (p$=$0.07) and only amounted to a 6\% average decrease (290 grains per stigma at the field edge vs. 240 at field centre).
% Flies were also common visitors in commodity fields (Table \ref{tab:propVisitors}), but had no effect on pollen deposition (p$=$0.577), so only honey bees and LCBs were considered as potential pollinators (other flower visitors were rare).
In seed canola, pollen deposition was much lower overall (mean: 22 grains per stigma, SD: 43, range: 0--578), and decreased from 23 to 11 grains/stigma from the edge to the centre of the field (Z=-4.4, p$=$0.0001, Figure \ref{fig:distCentPol}).
LCB visits increased pollen deposition from 12 grains/stigma at 6 visits/hr to 16 pollen grains at 60 visits/hr (Z=3.2, p$=$0.002, Figure \ref{fig:visPol_both}), but there was no effect of honey bee visitation (Z=0.98, p$=$0.33), implying that most of the pollen deposition in seed fields is the result of LCB visitation.
Pollen deposition also decreased from 23 grains/stigma at the edge of the female bay to 13 at the centre of the bay independent of visitation (Z=-4.5, p$=$0.0001), suggesting that pollen may be lost from bees crossing over from male bays \citep{pinnisch1990}.%, who likely visit flowers at the edge of the bay first.

\begin{figure}
    \centering
    \includegraphics[width=0.9\textwidth,keepaspectratio=true]{slopeHbeeVisPol.png}
    \caption[Partial effect of honey bee visitation on pollen deposition in commodity canola]{Partial effect of honey bee visitation on pollen deposition in commodity canola. Solid line represents median, and shaded areas represent 95\% CIs.}
    \label{fig:hbeeVisPol}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=0.9\textwidth,keepaspectratio=true]{slopeVisPol_both.png}
    \caption[Partial effects of honey bee and leafcutter bee visitation on pollen deposition in seed canola]{Partial effects of honey bee and leafcutter bee visitation on pollen deposition in seed canola. Solid lines represent medians, and shaded areas represent 95\% CIs.}
    \label{fig:visPol_both}
\end{figure}

\begin{figure} 
    \centering
    \includegraphics[width=0.9\textwidth,keepaspectratio=true]{slopeDistCentPol.png}
    \caption[Partial effect of distance from field edge and bay position on pollen deposition in seed canola]{Partial effect of distance from field edge and bay position on pollen deposition in seed canola. Coloured points represents median, and shaded areas represent 95\% CIs.}
    \label{fig:distCentPol}
\end{figure}

@@

\subsection{Prediction 4: Pollinated plants produce more seeds}

Pollen deposition increased fruit production and seed production in seed canola, but not in commodity canola.
Pollen deposition had no effect on pod set (p=0.12), seeds per pod (p=0.94) or seed size (p=0.56) in commodity canola, and there was also no direct effect of honey bee visitation on pod set (p=0.38), seeds per pod (p=0.55), or seed size (p=0.34).
Seed canola, in contrast, was highly dependent on pollen deposition.
Pollen deposition increased pod set (Z=3.1, p$=$0.002), and seeds per pod (Z=2.3, p$=$0.02), but had no direct effect on seed size (p$=$0.35).
Pod set also decreased with three types of distance: distance into the field (Z=-5.3, p$<$0.0001), distance from leafcutter shelters (Z=-3.2, p$=$0.002), and distance from the edge of the female bay (Z=-4.0, p$=$0.0001).
Seeds per pod decreased in the centre of the female bay (Z=-4.6, p$<$0.0001), and was higher in plants with high pod set (Z=6.0, p$<$0.0001).

@@

\subsection{Pod production} % The total yield of a canola plant is a function of three components: number of pods per plant, number of seeds per pod, and seed size.
Pod set in commodity canola was positively affected by plant size, but not by pollen deposition or honey bee visitation.
Both pod production and seeds per pod can increase with extra pollination \citep{jauker2008,sabbahi2005,sabbahi2006,duran2010}, so this lack of effect may be due to a high overall level of pollen deposition (Figure \ref{fig:hbeeVisPol}), even in fields without honey bees.
There was no effect of honey bee visitation on pod set, seeds per pod, or seed size, indicating that visitation did not seem to cause either \textit{a)} a greater quantity of pollen (see above) or \textit{b)} higher quality (outcrossed) pollen.
Self-pollination (geitanogamy) in canola can result in similar fruit production as outcrossing \citep{rosa2011}, implying that pollen quality is not likely a limiting factor in canola pod set, at least in some varieties.
Flower number was negatively related to pod set, which implies a response to pollination (high flower fertilization suppresses further flower production) and/or a reproductive trade-off (greater pod development rate means more competition among pods for resources).
%RC: THESE SCENARIOS AREN'T MUTUALLY EXCLUSIVE
We considered the first scenario to be the stronger effect, as \citet{sabbahi2006} and \citet{mesquida1981} both found that canola plants compensated for experimental removal of flowers by increasing branch and flower production, until the plant reached about 170 pods.
Their final number of pods is similar to the plants from our study (median: 112, range: 5--892), but we found that larger plants also had greater pod set, in addition to higher flower production, presumably caused by the larger pool of resources that large plants are able to invest into reproduction. 
%This could indicate stronger overall competition between plants in seed fields, but that plant resources are higher in the centre of the crop field than at the edge.

>>>>

Honey bee visitation decreased with distance away from the apiary, in both commodity and seed canola fields.
In commodity fields, honey bee visitation decreased from 6.4 visits/hr at the edge to 1.0 visits/hr at the centre of the field (400 m) (p$=$0.0002, Figure \ref{fig:allVis}, apiary of 40 hives).
% Visitation rates were higher in both years in Lethbridge (2014: 9.3 visits/hr, 2015: 7.1), but were higher in 2015 than 2014 at Grande Prairie (Z=2.3, p$=$0.02), rising from 5 visits/hr at the field edge in 2014 to 17 visits/hr in 2015. % (at a stocking level of 20 hives). 
Higher stocking rates increased honey bee visitation at the edge of the field from 3.9 visits/hr with an apiary of 20 hives, increasing to 6.4 visits/hr with 40 hives (Z=5.3, p$<$0.0001), while unstocked field edges had an average of 0.4 visits/hr.
Honey bee visitation was much higher overall in seed canola fields, and decreased from 99.5 visits/hr at edge to 78.8 visits/hr at the centre of the field (Z=-2.5, p$=$0.01). 
Honey bee visitation was also lower near leafcutter shelters (Z=3.9, p$=$0.001), dropping from 60 visits/hr at 10 m to 34 visits/hr at 2 m, likely due to competition with LCBs (Figure \ref{fig:hbeeDist_both}, \ref{fig:lbeeDist_both}).

@@

\begin{figure}
    \centering
    \includegraphics[width=0.9\textwidth,keepaspectratio=true]{../Figures/allVisits.png}
    \caption{Effect of distance from a) honey bee apiaries, b) leafcutter shelters, and c) female bay position on visitation rates. Commodity fields are shown at a stocking rate of 40 hives (black line), while stocking rates are at 160 hives in seed fields (yellow line).}
    \label{fig:allVis}
\end{figure}

@@

LCB visitation increased pollen deposition in seed canola fields, but honey bees did not increase pollen deposition in either commodity or seed canola fields.
Pollen deposition on stigmas was high in commodity canola (mean: 293 grains per stigma, SD: 385, range: 0--3981), but honey bee visitation had no impact on this (p$=$0.77, Figure \ref{fig:allPollen}).
Pollen deposition slighly declined with distance from the edge of the field, but this relationship was weak (p$=$0.07) and only amounted to a 6\% average decrease (290 grains per stigma at the field edge vs. 240 at field centre).
% Flies were also common visitors in commodity fields (Table \ref{tab:propVisitors}), but had no effect on pollen deposition (p$=$0.577), so only honey bees and LCBs were considered as potential pollinators (other flower visitors were rare).
In seed canola, pollen deposition was much lower overall (mean: 22 grains per stigma, SD: 43, range: 0--578), and decreased from 23 to 11 grains/stigma from the edge to the centre of the field (Z=-4.4, p$=$0.0001, Figure \ref{fig:distCentPol}).
LCB visits increased pollen deposition from 12 grains/stigma at 6 visits/hr to 16 pollen grains at 60 visits/hr (Z=3.2, p$=$0.002, Figure \ref{fig:visPol_both}), but there was no effect of honey bee visitation (Z=0.98, p$=$0.33), implying that most of the pollen deposition in seed fields is the result of LCB visitation.
Pollen deposition also decreased from 23 grains/stigma at the edge of the female bay to 13 at the centre of the bay independent of visitation (Z=-4.5, p$=$0.0001), suggesting that pollen may be lost from bees crossing over from male bays \citep{pinnisch1990}.%, who likely visit flowers at the edge of the bay first.

\begin{figure}
    \centering
    \includegraphics[width=0.9\textwidth,keepaspectratio=true]{allPollen.png}
    \caption{Effect of a) visitation rate, b) distance from apiary (HB) or shelter (LCB), and c) female bay position on pollen deposition.}
    \label{fig:allPollen}
\end{figure}

@@

% \subsection{Prediction 4: Pollinated plants produce more seeds}
% 
% Pollen deposition increased fruit production and seed production in seed canola, but not in commodity canola.
% Pollen deposition had no effect on pod set (p=0.12), seeds per pod (p=0.94) or seed size (p=0.56) in commodity canola, and there was also no direct effect of honey bee visitation on pod set (p=0.38), seeds per pod (p=0.55), or seed size (p=0.34).
% Seed canola, in contrast, was highly dependent on pollen deposition.
% Pollen deposition increased pod set (Z=3.1, p$=$0.002), and seeds per pod (Z=2.3, p$=$0.02), but had no direct effect on seed size (p$=$0.35).
% Pod set also decreased with three types of distance: distance into the field (Z=-5.3, p$<$0.0001), distance from leafcutter shelters (Z=-3.2, p$=$0.002), and distance from the edge of the female bay (Z=-4.0, p$=$0.0001).
% Seeds per pod decreased in the centre of the female bay (Z=-4.6, p$<$0.0001), and was higher in plants with high pod set (Z=6.0, p$<$0.0001).

@@

\subsection{Pod production} % The total yield of a canola plant is a function of three components: number of pods per plant, number of seeds per pod, and seed size.
Pod set in commodity canola was positively affected by plant size, but not by pollen deposition or honey bee visitation.
Both pod production and seeds per pod can increase with extra pollination \citep{jauker2008,sabbahi2005,sabbahi2006,duran2010}, so this lack of effect may be due to a high overall level of pollen deposition (Figure \ref{fig:allPollen}), even in fields without honey bees.
There was no effect of honey bee visitation on pod set, seeds per pod, or seed size, indicating that visitation did not seem to cause either \textit{a)} a greater quantity of pollen (see above) or \textit{b)} higher quality (outcrossed) pollen.
Self-pollination (geitanogamy) in canola can result in similar fruit production as outcrossing \citep{rosa2011}, implying that pollen quality is not likely a limiting factor in canola pod set, at least in some varieties.
Flower number was negatively related to pod set, which implies a response to pollination (high flower fertilization suppresses further flower production) and/or a reproductive trade-off (greater pod development rate means more competition among pods for resources).
%RC: THESE SCENARIOS AREN'T MUTUALLY EXCLUSIVE
We considered the first scenario to be the stronger effect, as \citet{sabbahi2006} and \citet{mesquida1981} both found that canola plants compensated for experimental removal of flowers by increasing branch and flower production, until the plant reached about 170 pods.
Their final number of pods is similar to the plants from our study (median: 112, range: 5--892), but we found that larger plants also had greater pod set, in addition to higher flower production, presumably caused by the larger pool of resources that large plants are able to invest into reproduction. 
%This could indicate stronger overall competition between plants in seed fields, but that plant resources are higher in the centre of the crop field than at the edge.
